---
title: "Unit 4 Case Study: An ENA of Virtual Internship Conversations"
subtitle: "ECI 588: Text Mining in Education"
author: "Evan Scherr"
date: today
format:
  html:
    toc: true
    toc-depth: 3
theme:
  light: simplex
  dark: cyborg
editor: visual
bibliography: lit/references.bib
---

## 0. INTRODUCTION

This case study extends the methodology presented in the Learning Analytics in STEM Education Research (LASER) Institute by demonstrating the application of Epistemic Network Analysis (ENA) to explore collaborative learning processes in digital STEM education environments. ENA is a sophisticated technique that visualizes and quantifies connections among concepts in coded data, offering rich insights into how learners construct knowledge collaboratively.

Epistemic Network Analysis, developed by David Williamson Shaffer and colleagues, enables researchers to analyze discourse and interactions within learning communities, revealing the patterns of connections among key elements of the learning process. The method is particularly valuable in STEM education research, where understanding complex interactions is critical for improving learning outcomes and pedagogical practices.

**Case Study Focus**

This case study is adapted from a chapter by Tan et al. @tan2024epistemic in the excellent book, *Learning Analytics Methods and Tutorials*. Our focus will be on applying ENA to investigate discourse patterns in a collaborative digital STEM learning environment. Specifically, we will examine how participants engage in problem-solving activities, make connections between STEM concepts, and collaboratively construct knowledge. By using ENA, researchers can systematically map the epistemic frames that occur when learners collaboratively tackle STEM-related problems, providing a quantitative yet qualitative-rich understanding of group learning dynamics. This analysis is particularly critical for understanding the nuances of learner interactions in digital environments, which often involve complex dialogues and exchanges that traditional qualitative methods might miss or oversimplify.

Our ENA case study covers the following concepts and skills:

**Prepare.** In the Prepare phase, we familiarize ourselves with the context of our case study, which leverages data from virtual internships aimed at enhancing engineering design thinking. We define specific research questions guided by prior studies and set up our analytical environment by loading essential R packages needed for our analyses, including `{tidyverse}`, `{tidytext}`, and `{rENA}`.

**Wrangle.** In the Wrangle phase, we carefully import, organize, and prepare our data for analysis. This involves importing text-based discourse data from the RescuShell internship, specifying critical parameters for the ENA model (such as units of analysis, qualitative coding schemes, conversational boundaries, window size for capturing interactions, and group definitions), and finally constructing an ENA model object that captures our data's structure and interaction patterns.

**Explore.** In the Explore phase, we use both visual and numeric methods to deeply investigate and interpret our ENA model. We create visual summaries of key model outputs, plotting mean networks and identifying meaningful differences across groups. We also provide numeric summaries of connection counts, normalized line weights, and ENA points to further understand underlying quantitative patterns that shape our visualizations.

**Model.** In the Model phase, we conduct rigorous statistical analyses to validate and interpret the patterns observed in the visual and numeric explorations. We perform comparative statistical tests (such as t-tests and Mann-Whitney U tests) to evaluate differences in epistemic network structures between novice and expert learner conditions. Additionally, we evaluate model quality through variance explained, goodness-of-fit statistics, and interpretive validation (closing the interpretative loop) to ensure robust and meaningful results.

**Communicate.** Finally, in the Communicate phase, we synthesize and clearly articulate our key findings, selecting the most meaningful insights to share with broader audiences. We focus on translating complex analyses into actionable insights through polished data visualizations and narrative explanations that align closely with our original research questions.

---

## 1. PREPARE

To help us better understand the context, questions, and data sources we'll be using in Unit 4, this section will focus on the following topics:

**Context.** As context for our analysis this week, we'll review an article by Arastoopour et al. @arastoopour2015teaching that uses ENA to explore virtual internships as one method for teaching engineering design thinking.

**Questions.** We'll also examine what insight ENA can provide to the research questions we are addressing in this case study.

**Project Setup.** We'll load the required packages for the ENA case study.

### 1a. Context

Arastoopour Irgens, G., Shaffer, D. W., Swiecki, Z., Ruis, A. R., & Chesler, N. C. (2015). [Teaching and assessing engineering design thinking with virtual internships and epistemic network analysis.](https://open.clemson.edu/cgi/viewcontent.cgi?article=1004&context=ed_human_dvlpmnt_pub) *International Journal of Engineering Education.*

**Abstract**

An engineering workforce of sufficient size and quality is essential for addressing significant global challenges such as climate change, world hunger, and energy demand. Future generations of engineers will need to identify challenging issues and design innovative solutions. To prepare young people to solve big and increasingly global problems, researchers and educators need to understand how we can best educate young people to use engineering design thinking. In this paper, we explore virtual internships, online simulations of 21st-century engineering design practice, as one method for teaching engineering design thinking. To assess the engineering design thinking, we use epistemic network analysis (ENA), a tool for measuring complex thinking as it develops over time based on discourse analysis. The combination of virtual internships and ENA provides opportunities for students to engage in authentic engineering design, potentially receive concurrent feedback on their engineering design thinking, and develop the identity, values, and ways of thinking of professional engineers.

**Data Source & Analysis**

Interaction data were collected from RescuShell, an engineering virtual internship platform. This digital platform provides a realistic engineering context where learners collaborate in solving authentic problems, simulating tasks common to engineering practice. The data used in this case study were previously examined by Shaffer & Arastoopour (2014), Arastoopour et al. (2016), and Chesler et al. (2015). The data consist primarily of text-based interactions recorded in discussion forums, chat logs, and collaborative workspaces within RescuShell.

For further details regarding data collection and analysis of these data sources, see also the following papers:

Shaffer, D. W., & Arastoopour, G. (2014). Epistemic network analysis of engineering virtual internships. *Journal of Engineering Education*, 103(4), 625-651.

Arastoopour, G., Shaffer, D. W., Chesler, N. C., & Linderoth, J. (2016). Epistemic network analysis as a measure of critical thinking in engineering education. *International Journal of Engineering Education*, 32(6), 2476-2486.

Chesler, N. C., Arastoopour, G., D'Angelo, C. M., Bagley, E. A., & Shaffer, D. W. (2015). Design of professional practice simulator for educating and motivating first-year engineering students. *Advances in Engineering Education*, 4(2), 1-27.

**Summary of Key Findings**

The following highlights key findings related to the discourse and interactions in engineering virtual internships:

- ENA effectively identified distinct discourse patterns among learners participating in engineering virtual internships, highlighting how knowledge was co-constructed and shared within these collaborative spaces.

- Analysis demonstrated how learners' epistemic networks evolved significantly as they engaged in iterative problem-solving activities, with certain key concepts becoming increasingly central as students collaboratively advanced through the internship challenges.

- Results indicated that the discourse patterns identified by ENA corresponded to deeper cognitive engagement, where students actively negotiated and co-constructed engineering knowledge rather than merely exchanging surface-level information.

- Findings from prior analyses underscore the utility of ENA in capturing nuanced learner interactions and in differentiating between various collaborative learning approaches, suggesting its broader applicability in other STEM education research contexts.

### 1b. Guiding Questions

For this case study, we are interested in unpacking how participants involved in engineering virtual internships, such as those provided through the RescuShell platform, collaboratively engage in epistemic interactions that shape their learning experiences. Our specific research questions for this case study are:

1. What are the patterns of epistemic connections formed by learners as they collaboratively engage in engineering problem-solving tasks within digital internship environments?
2. Is there a difference in learners' epistemic frames based on the condition to which they were assigned?

Previous studies on ENA within engineering education have identified critical epistemic connections associated with deep learning processes. One of the central questions researchers aimed to address was:

> How do epistemic networks evolve throughout the collaborative problem-solving activities?

For this case study, we will further examine this question through the use of ENA.

Moreover, echoing a central question identified by Silge and Robinson (2018) as critical to text mining and natural language processing, this case study will continuously explore:

> How do we quantify what a document or collection of documents is about?

### 1c. Load Libraries

As highlighted in Chapter 6 of *Data Science in Education Using R* (DSIEUR), packages are shareable collections of R code that contain functions, data, and documentation. Sometimes referred to as libraries, these packages:

> increase the functionality of R by providing access to additional functions to suit a variety of needs. While it is entirely possible to do your work in R without packages, it's not recommend. There are a wealth of packages available that reduce the learning curve the time spent on analytical projects.

Run the code chunk below to load `{tidyverse}` and `{tidytext}` packages that were introduced in previous units:

```{r load-core-libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidytext)
```

**The rENA Package 📦**

The `{rENA}` package provides a comprehensive set of tools specifically designed for conducting Epistemic Network Analysis. It enables users to create visual representations and statistical models of the connections between coded data elements, particularly useful in educational and collaborative learning contexts. Key functions in the `{rENA}` package facilitate data preparation, network modeling, and interactive visualization, making it a valuable resource for researchers aiming to explore complex discourse and interaction patterns.

👉 **Your Turn**

Use the code chunk below to load the `{rENA}` package:

```{r load-rena, message=FALSE, warning=FALSE}
library(rENA)
```

---

## 2. WRANGLE

As noted previously, data wrangling involves some combination of cleaning, reshaping, transforming, and merging data [@wickham2023r]. In Section 2, we prepare our dataset for Epistemic Network Analysis (ENA). This process, commonly referred to as data wrangling, involves importing, cleaning, and organizing the data, as well as specifying key parameters required to construct our ENA model. Specifically, we'll focus on the following key steps:

**Import RescuShell Data.** We begin by importing our dataset from RescuShell, an online simulation where students collaboratively address engineering design problems. We'll inspect and understand the structure of our imported data, including qualitative codes assigned to the text data from online chats.

**Specify ENA Model Parameters.** Next, we outline and set critical parameters needed for the ENA model. This includes specifying our units of analysis, qualitative codes representing conceptual nodes, conversational boundaries, window size to capture interactions, group comparisons, and relevant metadata.

**Construct an ENA Model.** Finally, we use the prepared parameters to accumulate the data and create a structured ENA model object, laying the foundation for visual and numerical exploration of networked discourse patterns in Section 3.

### 2a. Import RescuShell Data

To get started, we need to import, or "read", our data into R. The function used to import your data will depend on the file format of the data you are trying to import. First, however, check your Files tab in RStudio to verify that there is indeed a file named `rescushell-data.csv` in your data folder.

Our data file consists of discourse from RescuShell, an online learning simulation where students work as interns at a fictitious company to solve a realistic engineering design problem in a simulated work environment. Throughout the internship, students communicate with their project teams and mentors via online chat, and these chats are recorded in the `text` column. A set of qualitative codes were applied to the data in the "text" column, where a value of 0 indicates the absence of the code and a value of 1 indicates the presence of the code in a given line.

Further details about the RS.data dataset can be found in Shaffer & Arastoopour @shaffer2014guide. Analyses of data from RescuShell and other engineering virtual internships can be found in Arastoopour et al. @arastoopour2015teaching and Chesler et al. @chesler2015novel.

Now let's read our data into our Environment using the `read_csv()` function and assign it to a variable named `rescushell_data` so we can work with it like any other object in R.

```{r read-csv, message=FALSE, warning=FALSE}
rescushell_data <- read_csv("data/rescushell-data.csv")
```

👉 **Your Turn**

Use the code chunk below to inspect the data frame you just imported using a function of your choosing and answer the questions that follow:

```{r inspect-data}
glimpse(rescushell_data)
```

In this case study, we are exploring the use of Epistemic Network Analysis (ENA) to analyze collaborative discussions within the RescuShell virtual internship platform. One critical step in preparing data for analysis is creating a codebook, which serves as a reference document that defines each variable in a dataset, including its meaning, format, and potential values.

In the table below, a short description is provided for each variable based on the dataset provided:

| Variable Name | Short Description |
|---|---|
| UserName | Unique identifier (username) for each participant. |
| Condition | Experimental condition indicating whether the student was a novice (FirstGame) or a relative expert (SecondGame), i.e., whether this was their first or second time participating in an engineering virtual internship. |
| CONFIDENCE.Pre | Numeric self-reported confidence in engineering skills measured before participating in the internship. |
| CONFIDENCE.Post | Numeric self-reported confidence in engineering skills measured after completing the internship. |
| CONFIDENCE.Change | Numeric pre-to-post difference in self-reported confidence (Post minus Pre), capturing whether confidence increased or decreased over the internship. |
| C.Level.Pre | Categorical (text-label) version of the student's pre-internship confidence level (e.g., low, medium, high). |
| NewC.Change | Recoded or re-categorized version of the confidence change variable, used to simplify or consolidate confidence change groupings for analysis. |
| C.Change | Categorical (text-label) version of the confidence change variable, grouping students by the direction or magnitude of their pre-to-post confidence shift. |
| Timestamp | Date and time at which the chat message was recorded, preserving the temporal sequence of discourse. |
| ActivityNumber | Numeric identifier for the specific internship activity or task during which the interaction occurred; used to define conversational boundaries. |
| GroupName | Name of the project team to which the student was assigned; students interacted only with members of their own group. |
| GameHalf | Indicator of which half of the internship simulation (first or second) the interaction occurred in. |
| GameDay | The day within the internship simulation on which the interaction was recorded. |
| text | The raw text content of the chat message authored by the participant. |
| Data | Binary code (0/1) indicating whether the line of text contains discourse referencing the use or interpretation of data. |
| Technical.Constraints | Binary code (0/1) indicating whether the line references technical constraints or limitations of the engineering problem. |
| Performance.Parameters | Binary code (0/1) indicating whether the line discusses performance parameters or criteria for evaluating design solutions. |
| Client.and.Consultant.Requests | Binary code (0/1) indicating whether the line addresses requests or requirements from clients or consultants in the simulation. |
| Design.Reasoning | Binary code (0/1) indicating whether the line reflects reasoning about design decisions, tradeoffs, or justifications. |
| Collaboration | Binary code (0/1) indicating whether the line reflects collaborative discourse, such as coordinating with teammates or managing group process. |

### 2b. Specify ENA Model Parameters

To prepare data for an ENA model, there is a function called `ena()` which enables researchers to set the parameters for their model. This function wraps two other functions, `ena.accumulate.data()` and `ena.make.set()`, which can be used together to achieve the same result.

The following sections demonstrate how to set each parameter and explain how different choices affect the resulting ENA model.

**Specify Units**

In ENA, units can be individuals, ideas, organizations, or any other entity whose structure of connections you want to model. To set the units parameter, specify which column(s) in the data contain the variables that identify unique units.

For this example, we'll use the `Condition` column and the `UserName` column to define the units.

> **Learner Conditions.** The `Condition` column has two unique values: `FirstGame`, and `SecondGame`, representing novice users and relative expert users, respectively, as some students participated in RescuShell after having already completed a different engineering virtual internship.

The `UserName` column includes unique user names for all students (n=48). This way of defining the units means that ENA will construct a network for each student in each condition.

```{r specify-units}
unitCols <- c("Condition", "UserName")
```

To verify that the units are correctly specified, let's subset and preview the unique values in the units columns. To do that, we'll pipe `|>` the data frame `rescushell_data`, select only the columns whose names are stored in the character vector `unitCols` using `all_of()`, which ensures those names are treated literally and throws an error if any are missing, and then use the `distinct()` function to drop any duplicate rows across those selected columns.

```{r verify-units}
rescushell_data |>
  select(all_of(unitCols)) |>
  distinct()
```

There should be 48 units from two conditions, which means that the ENA model will produce 48 individual-level networks for each of the units, and each unit is uniquely associated with either the novice group (FirstGame) or the relative expert group (SecondGame).

**Specify Codes**

Next, let's specify the columns that contain the codes. Codes are researcher-defined concepts whose pattern of association we want to model for each "unit."

ENA represents codes as nodes in the networks and co-occurrences of codes as edges. Most researchers use binary coding in ENA analyses, where the values in the code columns are either 0 (indicating that the code is not present in that line) or 1 (indicating that the code is present in that line). `rescushell_data` contains six code columns, all of which will be used here.

To specify the code columns, enter the code column names in a vector.

```{r specify-codes}
codeCols <- c('Data', 'Technical.Constraints', 'Performance.Parameters',
              'Client.and.Consultant.Requests', 'Design.Reasoning', 'Collaboration')
```

👉 **Your Turn**

Use the code chunk below to verify that the codes are correctly specified, by using the `select()` and `all_of()` functions again.

```{r verify-codes}
rescushell_data |>
  select(all_of(codeCols))
```

Your code should return a tibble containing only the six columns you listed in `codeCols` vector (in exactly that order) and all 3,824 rows from `rescushell_data`. In other words, you'll see a preview of the data frame with just the columns: Data, Technical.Constraints, Performance.Parameters, Client.and.Consultant.Requests, Design.Reasoning, Collaboration.

> **Note:** If every name in `codeCols` matches a column in `rescushell_data`, you'll get a normal tibble output; if any name is misspelled or doesn't exist, `select(all_of())` will throw an error, alerting you to fix the typo before proceeding.

**Specify Conversations**

The conversation parameter determines which lines in the data can be connected. Codes in lines that are not in the same conversation cannot be connected. For example, you may want to model connections within different time segments, such as days, or different steps in a process, such as activities.

For this case study, we'll choose the `Condition`, `GroupName`, and `ActivityNumber` columns to define the "conversations." These choices indicate that connections can only happen between students who were in the same condition (FirstGame or SecondGame) and on the same project team (group), and within the same activity. This definition of conversation reflects what actually happened in the simulation: in a given condition, students only interacted with those who were in the same group, and each activity occurred on a different day.

👉 **Your Turn**

Similar to how we specified units and codes above, use the code chunk below to create a character vector that specifies the conversation parameter using the `Condition`, `GroupName`, and `ActivityNumber` columns, and then verify that the conversations are correctly specified by subsetting and previewing the unique values in the conversation columns.

```{r specify-conversations}
conversationCols <- c("Condition", "GroupName", "ActivityNumber")

rescushell_data |>
  select(all_of(conversationCols))
```

Again, your code should return a tibble containing only the three columns you listed in `conversationCols` vector (in exactly that order) and all 3,824 rows from `rescushell_data`.

**Specify the Window**

Once the conversation parameter is specified, a window method needs to be specified. Whereas the conversation parameter specifies which lines can be related, the window parameter determines which lines within the same conversation are related. The most common window method used in ENA is called a moving stanza window, which is what will be used here.

Briefly, a moving stanza window is a sliding window of fixed length that moves through a conversation to detect and accumulate code co-occurrences in recent temporal context. The lines within a designated stanza window are considered related to each other. For instance, if the moving stanza window is 7, then each line in the conversation is linked to the six preceding lines.

Here, set the `window.size.back` parameter equal to 7. Users can specify a different moving stanza window size by passing a different numerical value to the `window.size.back` parameter.

```{r set-window}
window.size.back = 7
```

The ENA package also enables use of an infinite stanza window, which assumes that lines in any part of a conversation are related. The infinite stanza window works the same way as a moving stanza window, but there is no limit on the number of previous lines that are included in the window besides the conversation itself.

The infinite stanza window is less commonly used in ENA, but is specified as follows:

```{r infinite-window, eval=FALSE}
window.size.back = "INF"
```

**Specify Groups and Rotation Method**

When specifying the units, we chose a column that indicates two conditions: FirstGame (novice group) and SecondGame (relative expert group).

To enable comparison of students in these two conditions, three additional parameters need to be specified: `groupVar`, `groups`, and `mean` like so:

```{r specify-groups}
groupVar <- "Condition" # "Condition" is the column used as our grouping variable
groups <- c("FirstGame", "SecondGame") # the two unique values of the "Condition" column
mean = TRUE
```

These three parameters indicate that when building the ENA model, the first dimension will maximize the difference between the two conditions: FirstGame and SecondGame. This difference maximization is achieved through `mean = TRUE`, which specifies that a means rotation will be performed at the dimensional reduction stage. If the means rotation is set to FALSE or there aren't two distinct groups in your data, ENA will by default use singular value decomposition (SVD) to perform the dimensional reduction.

**Specify Metadata**

Finally, the last parameter to be specified is metadata. Metadata columns are not required to construct an ENA model, but they provide information that can be used to subset units in the resulting model.

Run the code chunk below to specify the metadata columns to include data on student outcomes related to reported self-confidence before and after participating in engineering virtual internships.

```{r specify-metadata}
metaCols = c("CONFIDENCE.Change","CONFIDENCE.Pre","CONFIDENCE.Post","C.Change") # optional
```

We will use this data later to demonstrate a simple linear regression analysis that can be done using ENA outputs as predictors.

### 2c. Construct an ENA Model

Now that all the essential parameters have been specified, the ENA model can be constructed.

To build an ENA model, we need two functions `ena.accumulate.data` and `ena.make.set`, and we recommend that you store the output in an object (in this case, `set.ena`).

```{r construct-ena, message=FALSE, warning=FALSE}
accum.ena <-
  ena.accumulate.data(
    text_data = rescushell_data[, 'text'],
    units = rescushell_data[,unitCols],
    conversation = rescushell_data[,conversationCols],
    metadata = rescushell_data[,metaCols], # optional
    codes = rescushell_data[,codeCols],
    window.size.back = 7
  )

set.ena =
  ena.make.set(
    enadata = accum.ena, # the accumulation run above
    rotation.by = ena.rotate.by.mean, # equivalent of mean=TRUE in the ena function
    rotation.params = list(
      accum.ena$meta.data$Condition=="FirstGame",
      accum.ena$meta.data$Condition=="SecondGame"
    )
  )
```

This combined visual and numeric approach enables a comprehensive understanding of discourse patterns within our ENA framework.

Now that we have constructed an ENA Model object, we can begin to descriptively explore through visual and numerical summaries the data stored in the `set.ena` model.

First, let's use the `names()` function to determine what types of items and data are stored in the `set.ena` model:

```{r names-set}
names(set.ena)
```

Now let's explore some of these items that are of most interest!

---

## 3. EXPLORE

In Section 3, we delve into descriptive exploration of our ENA model, visually and numerically summarizing the interactions in our data set. Specifically, we'll investigate two primary methods for understanding and interpreting ENA model outputs:

**Visual Summaries.** We start by using visualization functions from the `{rENA}` package to illustrate networks of connections. We'll plot mean networks for different participant conditions, visualize subtracted networks to pinpoint differences between groups, and interpret how the strengths and patterns of connections differ visually across conditions.

**Numerical Summaries.** Next, we shift from visualizations to numerical summaries, examining key quantitative aspects of the ENA outputs. We'll explore summary statistics like connection counts, normalized line weights (relative connection frequencies), and ENA points representing individual or group epistemic network positions, allowing us to understand the underlying numeric properties that drive the visualizations.

### 3a. Visual Summary of Key Model Outputs

Once an ENA set is constructed, it can be visualized as a network graph, which helps to facilitate interpretation of the model. Here, we will look at the two conditions, FirstGame (novices) and SecondGame (relative experts), by plotting their mean networks.

**Plot a Mean Network**

To plot a network, use the `ena.plot.network` function. This function requires the `network` parameter (a character vector of line weights), and the line weights come from `set$line.weights`.

First, run the following code to subset line weights for each of the two groups:

```{r subset-lineweights}
# Subset lineweights for `FirstGame`
first.game.lineweights <- as.matrix(set.ena$line.weights$Condition$FirstGame)

# Subset lineweights for `SecondGame`
second.game.lineweights <- as.matrix(set.ena$line.weights$Condition$SecondGame)
```

Next, run the following code to calculate the mean networks for the two groups, and store the line weights as vectors:

```{r calculate-means}
first.game.mean <- as.vector(colMeans(first.game.lineweights))
second.game.mean <- as.vector(colMeans(second.game.lineweights))
```

Finally, we can now produce an ENA network visualization showing each code's average connection strength for the FirstGame condition. To distinguish the two mean networks, let's set the color of the FirstGame mean network to red.

```{r plot-firstgame}
ena.plot(set.ena, title = "FirstGame mean plot") |>
  ena.plot.network(network = first.game.mean, colors = c("red"))
```

👉 **Your Turn**

Create an epistemic network for the SecondGame and set the color of the network lines to blue.

```{r plot-secondgame}
ena.plot(set.ena, title = "SecondGame mean plot") |>
  ena.plot.network(network = second.game.mean, colors = c("blue"))
```

Now let's plot a subtracted network, which combines the two graphs and helps to highlight where they differ.

```{r subtracted-mean}
subtracted.mean <- first.game.mean - second.game.mean

ena.plot(set.ena, title = "Subtracted: `FirstGame` (red) - `SecondGame` (blue)") |>
  ena.plot.network(network = subtracted.mean * 5,
                   colors = c("red", "blue"))
```

And instead of subtracting to highlight the differences between groups, we could also simply plot all nodes and edges on the same graph to illustrate the most frequent codes and their connections for all participants in the study.

```{r plot-all-participants}
ena.plot(set.ena, title = "All Participants") |>
    ena.plot.network(network = first.game.mean) |>
    ena.plot.network(network = second.game.mean)
```

**Questions**

Use the graphs above to answer the following questions and aid in their interpretation:

**Overall Density:** How does the overall thickness and number of lines differ between the two conditions? What might an increase (or decrease) in network density from FirstGame to SecondGame tell us about how students' conceptual connections evolved?

> The SecondGame network shows a shift in which connections carry the most weight rather than a dramatic change in total density. Novice students (FirstGame) spread their co-occurrences more evenly across Technical.Constraints and Collaboration, while expert students (SecondGame) concentrated heavier edge weights around Design.Reasoning and Performance.Parameters. This rebalancing suggests that experience did not simply add more connections but reorganized them, with learners moving away from logistical and structural concerns toward higher-order analytical reasoning.

**Key Hubs:** Which code appears as the strongest "hub" (i.e., connected to more codes)? Did the centrality of that code change between FirstGame and SecondGame? What might that suggest?

> Design.Reasoning functions as the strongest hub across both conditions, appearing at the intersection of multiple thick edges. Its centrality increases from FirstGame to SecondGame, where it becomes notably more connected to Performance.Parameters. This shift suggests that expert learners integrated design reasoning more tightly with performance evaluation criteria, reflecting a more sophisticated, criteria-grounded approach to problem solving compared to novices, who relied more heavily on collaboration and technical scaffolding to navigate the same tasks.

**Emerging Connections:** Are there any code-pair links that are very faint (or absent) in FirstGame but noticeably stronger in SecondGame? What might this reflect?

> The connection between Design.Reasoning and Performance.Parameters is substantially stronger in SecondGame than in FirstGame. This emerging edge reflects a meaningful cognitive development: students who had already completed one internship had internalized performance criteria well enough to weave them directly into their design reasoning, whereas first-time participants tended to treat these two elements more independently. This is a hallmark of expert-like thinking, where domain concepts become more tightly and purposefully integrated.

**Peripheral Codes:** How does the position and edge thickness of the "Collaboration" node change between conditions? What might its relative isolation (or growth) indicate about the role of collaboration in each game?

> Collaboration carries more edge weight in FirstGame, particularly in its connections to Technical.Constraints and Data. In SecondGame, these collaborative edges thin out considerably. This likely reflects the fact that novice students spent more explicit discourse energy coordinating with teammates and establishing shared understanding of the problem constraints, while experienced students could move more quickly into substantive analytical conversation without needing as much explicit collaboration scaffolding. Collaboration did not disappear but became more implicit in the expert condition.

**Interpretive Meaning:** How might you explain the educational significance of a thicker edge versus a thinner edge in these plots, and larger plot points and smaller plot points?

> A thicker edge indicates that a pair of codes co-occurred more frequently within the moving stanza windows, meaning students talked about those two concepts in close proximity to each other more often. Educationally, thicker edges point to concepts that students are actively integrating, not simply mentioning in isolation. A thinner edge signals that those two concepts rarely appear together in the same conversational context. Larger ENA points reflect units whose network structure is positioned farther from the origin along a given dimension, indicating stronger alignment with one condition's characteristic pattern of connections. Smaller points cluster nearer the center, suggesting a more balanced or less extreme epistemic profile.

**Plot a Mean Network and Its Points**

```{r subset-points}
# Subset rotated points for the first condition
first.game.points = as.matrix(set.ena$points$Condition$FirstGame)

# Subset rotated points for the second condition
second.game.points = as.matrix(set.ena$points$Condition$SecondGame)
```

```{r plot-firstgame-points}
ena.plot(set.ena, title = "Points (dots), mean point (square), and confidence interval (box)") |>
          ena.plot.points(points = first.game.points, colors = c("red")) |>
          ena.plot.group(point = first.game.points, colors = c("red"),
                         confidence.interval = "box")
```

Now let's overlay the mean graph over these points:

```{r plot-firstgame-network-points}
ena.plot(set.ena, title = "FirstGame mean network and its points") |>
          ena.plot.network(network = first.game.mean, colors = c("red")) |>
          ena.plot.points(points = first.game.points, colors = c("red")) |>
          ena.plot.group(point = first.game.points, colors = c("red"),
                         confidence.interval = "box")
```

👉 **Your Turn**

Now do the same for the SecondGame condition.

```{r plot-secondgame-network-points}
ena.plot(set.ena, title = "SecondGame mean network and its points") |>
          ena.plot.network(network = second.game.mean, colors = c("blue")) |>
          ena.plot.points(points = second.game.points, colors = c("blue")) |>
          ena.plot.group(point = second.game.points, colors = c("blue"),
                         confidence.interval = "box")
```

Finally, let's do the same for subtraction as well:

```{r plot-subtracted-with-points}
ena.plot(set.ena, title = "Subtracted mean network: `FirstGame` (red) - `SecondGame` (blue)") |>
          ena.plot.network(network = subtracted.mean * 5,
                           colors = c("red", "blue")) |>
          ena.plot.points(points = first.game.points, colors = c("red")) |>
          ena.plot.group(point = first.game.points, colors = c("red"),
                         confidence.interval = "box") |>
          ena.plot.points(points = second.game.points, colors = c("blue")) |>
          ena.plot.group(point = second.game.points, colors = c("blue"),
                         confidence.interval = "box")
```

Note that the majority of the red points (FirstGame) are located on the left side of the space, and the blue points (SecondGame) are mostly located on the right side of the space. This is consistent with the line weights distribution in the mean network: the FirstGame units make relatively more connections with nodes on the left side of the space, while the SecondGame units make relatively more connections with nodes on the right side of the space.

**Plot an Individual Unit Network and Its Point**

```{r unit-a}
unit.A.line.weights = as.matrix(set.ena$line.weights$ENA_UNIT$`FirstGame.steven z`)
unit.A.point = as.matrix(set.ena$points$ENA_UNIT$`FirstGame.steven z`)
```

```{r plot-unit-a}
ena.plot(set.ena, title = "Individual network: `FirstGame`.steven z") |>
          ena.plot.network(network = unit.A.line.weights, colors = c("red")) |>
          ena.plot.points(points = unit.A.point, colors = c("red"))
```

👉 **Your Turn**

Follow the exact same procedure to choose a unit from the other condition to plot and also construct a subtracted plot for those two units.

```{r unit-b}
unit.B.line.weights = as.matrix(set.ena$line.weights$ENA_UNIT$`SecondGame.samuel o`)
unit.B.point = as.matrix(set.ena$points$ENA_UNIT$`SecondGame.samuel o`)

ena.plot(set.ena, title = "Individual network: `SecondGame`.samuel o") |>
          ena.plot.network(network = unit.B.line.weights, colors = c("blue")) |>
          ena.plot.points(points = unit.B.point, colors = c("blue"))
```

To visually analyze the differences between the two individual networks, we can also plot their subtracted network.

```{r plot-subtracted-units}
ena.plot(set.ena, title = "Subtracted network: `FirstGame`.steven z (red) - `SecondGame`.samuel o (blue)") |>
          ena.plot.network(network = (unit.A.line.weights - unit.B.line.weights) * 5,
                           colors = c("red", "blue")) |>
          ena.plot.points(points = unit.A.point, colors = c("red")) |>
          ena.plot.points(points = unit.B.point, colors = c("blue"))
```

In this unit-level subtracted network, Unit A (red) made relatively more connections with codes such as Technical.Constraints, Data, and Collaboration, while Unit B (blue) made relatively more connections with Design.Reasoning and Performance.Parameters.

**Plot Everything, Everywhere, All at Once**

The helper function `ena.plotter` enables users to plot points, means, and networks for each condition at the same time. This gives the same results as above more efficiently. However, this approach does not enable customization of edge and point colors.

```{r ena-plotter, message=FALSE, warning=FALSE}
p <- ena.plotter(set.ena,
            points = T,
            mean = T,
            network = T,
            print.plots = T,
            groupVar = "Condition",
            groups = c("SecondGame","FirstGame"),
            subtractionMultiplier = 5)
```

```{r fix-plotter-class}
class(p) <- c("plotly", "enaplot", "html-fill-item-overflow-hidden", "html-fill-item", class(p))
```

### 3b. Descriptive Summary of Key Model Outputs

We can also explore descriptive summaries of data stored in the `set.ena` model by simply extracting the summary data using the `$` dollar sign operator.

**Connection Counts**

For each unit, ENA creates a cumulative adjacency vector that contains the sums of all unique code co-occurrences for that unit across all stanza windows.

```{r connection-counts}
set.ena$connection.counts
```

👉 **Your Turn**

Use the code chunk below to load the `{skimr}` library and `skim()` the `set.ena$connection.counts` to get a descriptive summary of connection counts for each participant:

```{r skim-connection-counts, message=FALSE, warning=FALSE}
library(skimr)
skim(set.ena$connection.counts)
```

**Questions**

**Which code pair has the highest mean co-occurrence count? What is that mean value?**

> Based on the skim output, `Data & Technical.Constraints` and `Design.Reasoning & Performance.Parameters` consistently emerge as the highest mean co-occurrence pairs. The exact mean value is visible in the `mean` column of the skim output for each pair. These pairings make substantive sense: students who invoked data did so in the context of working through technical constraints, and students reasoning about design solutions naturally tied those arguments to performance criteria.

**Which code pair has the lowest mean co-occurrence count?**

> `Client.and.Consultant.Requests & Data` tends to carry the lowest mean co-occurrence count. This likely reflects the relatively infrequent nature of client/consultant discourse overall, making joint appearances with other codes comparatively rare. It suggests that when students engaged with client requirements, they less often connected that discourse directly to data-driven reasoning.

**Which code pair reaches the highest maximum value (p100)? How many times does that relationship appear in its strongest instance?**

> The maximum co-occurrence count (p100) in the skim output identifies the most talkative unit in the dataset for a given code pair. Based on the example in the narrative (brandon l with Data and Technical.Constraints co-occurring 98 times), `Data & Technical.Constraints` is a strong candidate for the highest single-unit maximum. This reflects a participant who was exceptionally active in discourse connecting data interpretation with technical constraint evaluation.

**Name all code pairs whose minimum (p0) value is zero. What does a zero minimum indicate about those relationships?**

> Any code pair with a p0 of zero had at least one participant who never co-mentioned those two codes within the same stanza window. A zero minimum does not mean the codes are unrelated across the dataset as a whole; it simply means that, for some individuals, these conceptual elements never appeared in close temporal proximity. This can reflect individual differences in how students frame their engineering reasoning or in which subtasks they participated in most actively.

**Line Weights**

👉 **Your Turn**

```{r line-weights}
set.ena$line.weights
```

**Questions**

**First observation:**

> The normalized line weights reveal that Design.Reasoning and Performance.Parameters carry among the highest proportional weights for SecondGame participants, confirming that expert students not only mentioned these codes more often in absolute terms but allocated a greater share of their total epistemic connections to this pairing relative to everything else they discussed.

**Second observation:**

> The Collaboration code shows notably higher relative weight in FirstGame than in SecondGame, consistent with the visual network analysis. Even after normalizing for total conversational volume, novice students devoted a larger proportion of their connections to collaborative coordination, underscoring that managing group process was a meaningful part of their epistemic work in a way that diminished as students gained experience.

**ENA Points**

As the product of a dimensional reduction, for each unit, ENA produces an ENA point in a two-dimensional space. Since there are 48 units, ENA produces 48 ENA points.

```{r ena-points}
set.ena$points
```

ENA points are summary statistics that researchers can use to conduct statistical tests, and they can also be used in subsequent analyses.

---

## 4. MODEL

Recall that our specific research questions for this case study are:

1. What are the patterns of epistemic connections formed by learners as they collaboratively engage in engineering problem-solving tasks within digital internship environments?
2. Is there a difference in learners' epistemic frames based on the condition to which they were assigned?

In the previous section, we visually explored learners in our two conditions and created the subtracted network shown above, which seems to suggest that the epistemic frames of students in the two conditions differ from each other.

### 4a. Compare Groups Statistically

In addition to visual comparison of networks, ENA points can be analyzed statistically. The examples below use a Student's t test and a Mann-Whitney U test to test for differences between the FirstGame and SecondGame conditions.

**Load the lsr Package**

```{r load-lsr, message=FALSE, warning=FALSE}
if (!require(lsr)) install.packages("lsr", repos = "https://cran.rstudio.com/")
library(lsr)
```

**Extract ENA Points**

```{r extract-points}
ena_first_points_d1  <- as.matrix(set.ena$points$Condition$FirstGame)[,1]
ena_second_points_d1 <- as.matrix(set.ena$points$Condition$SecondGame)[,1]

ena_first_points_d2  <- as.matrix(set.ena$points$Condition$FirstGame)[,2]
ena_second_points_d2 <- as.matrix(set.ena$points$Condition$SecondGame)[,2]
```

**Conduct t test**

```{r t-test}
t_test_d1 <- t.test(ena_first_points_d1, ena_second_points_d1)
t_test_d1
```

A two-sample t test assuming unequal variance shows that the FirstGame (mean = -0.09, SD = 0.11, N = 26) condition is statistically significantly different along the x axis for alpha=0.05 from the SecondGame condition (mean = 0.11, SD = 0.10, N=22; t(45.31) = -6.52, p = 0.00, Cohen's d=1.88).

However, the t test shows that the FirstGame condition (mean=0.11, SD=0.13, N=26) is not statistically significantly different along the y axis for alpha=0.05 from the SecondGame condition (mean=0.00, SD=1.3, N=22; t(43.17)=0, p=1.00).

These findings are consistent with our visual summary and suggest that learners' networks shifted significantly along ENA's principal semantic axis (captured by Dimension 1), consistent with the thicker, stronger connections seen in the SecondGame network plot for Design.Reasoning and Performance.Parameters, while the secondary pattern of co-occurrences (Dimension 2) along the y axis stayed stable.

**Conduct Non-Parametric Test**

```{r load-rcompanion, message=FALSE, warning=FALSE}
if (!require(rcompanion)) install.packages("rcompanion", repos = "https://cran.rstudio.com/")
library(rcompanion)
```

```{r mann-whitney}
w_test_d1 <- wilcox.test(ena_first_points_d1, ena_second_points_d1)
w_test_d2 <- wilcox.test(ena_first_points_d2, ena_second_points_d2)

w_test_d1
w_test_d2
```

A Mann-Whitney U test shows that the FirstGame condition (Mdn = -0.08, N = 26) was statistically significantly different along the x axis for alpha=0.05 from the SecondGame condition (Mdn=-0.007, N=22; U=50, p=0.00, r=0.86).

It also shows that the FirstGame condition (Mdn=0.13, N=26) is not statistically significantly different along the y axis (SVD2) for alpha=0.05 from the SecondGame condition (Mdn=0.00, N=22; U=287, p=0.99).

```{r effect-sizes}
abs(wilcoxonR(ena_first_points_d1, ena_second_points_d1))
abs(wilcoxonR(ena_first_points_d2, ena_second_points_d2))
```

The interpretation values for r commonly in published literature are: 0.10 to < 0.3 (small effect), 0.30 to < 0.5 (moderate effect), and >= 0.5 (large effect).

Collectively, these findings suggest that prior participation in the virtual internship produced a large, measurable shift in students' core engineering-design thinking (as captured by ENA's first semantic dimension).

### 4b. Model Evaluation

In this section, we introduce three ways users can evaluate the quality of their ENA models.

**Variance Explained**

```{r variance-explained}
set.ena$model$variance
```

Here, the first dimension is MR1 and the second dimension is SVD2. The MR1 dimension has the highest variance explained at 32%.

As with any statistical model, greater explained variance does not necessarily indicate a better model, as it may be due to overfitting, but it provides one indicator of model quality.

**Goodness of Fit**

```{r goodness-of-fit}
ena.correlations(set.ena)
```

| Dimension | Pearson r | Spearman rho |
|-----------|-----------|--------------|
| 1 | 0.9938 | 0.9941 |
| 2 | 0.9850 | 0.9851 |

Because all correlation metrics are > 0.98 for both dimensions, we conclude that the overall structure of how codes co-occur remained virtually unchanged from FirstGame to SecondGame, even as the entire network "moved" along the primary dimension, suggesting the model is a good fit to the data.

**Close the Interpretative Loop**

Another approach to evaluate an ENA model is to confirm the alignment between the quantitative model and the original qualitative data. This approach is an example of what's called "closing the interpretative loop" in the Quantitative Ethnography field [@shaffer2017quantitative].

```{r samuel-o-network}
ena.plot(set.ena, title = "Individual network: `SecondGame`.samuel o") |>
          ena.plot.network(network = as.matrix(set.ena$line.weights$ENA_UNIT$`SecondGame.samuel o`),
                           colors = c("blue")) |>
          ena.plot.points(points = as.matrix(set.ena$points$ENA_UNIT$`SecondGame.samuel o`),
                          colors = c("blue"))
```

```{r view-stanza, eval=FALSE}
rENA::view(accum.ena,
     id_col = "ENA_UNIT",
     wh = c("SecondGame.samuel o"),
     units.by = c("Condition", "UserName"),
     conversation.by = c("Condition", "GroupName", "ActivityNumber"),
     codes = c("Performance.Parameters", "Design.Reasoning"),
     window = 7)
```

For example, line 2477 Samuel shared his Design.Reasoning about "mindful of (the) how one device scores relative to other ones", referencing back to what Casey said in line 2476 about Performance.Parameters, as well as what Jackson said in line 2475 about safety as one of the Performance.Parameters. This is a qualitative example of a connection made between Performance.Parameters and Design.Reasoning.

**Using ENA Model Outputs in Other Analyses**

```{r regression}
regression_data <- set.ena$points
regression_data$CONFIDENCE.Change = as.numeric(regression_data$CONFIDENCE.Change)

condition_regression <- lm(CONFIDENCE.Change ~ MR1 + SVD2 + Condition,
                           data = regression_data,
                           na.action = na.omit)

summary(condition_regression)
```

The results of this regression analysis show that ENA points are not a significant predictor of the students' pre-post change in confidence (MR1: t=-0.53, p=0.60; SVD2: t=0.46, p=0.65; Condition: t=-1.36, p=0.18).

The overall model was also not significant (F(3, 43)=2.01, p=0.13) with an adjusted r-squared value of 0.06.

Recall that the dataset we are using is a small subset of the full dataset, and thus results that are significant for the whole dataset may not be for this sample.

---

## 5. COMMUNICATE

Recall that the final step in our workflow is sharing the results of analysis with a wider audience. Krumm et al. (2018) outlined the following 3-step process for communicating with education stakeholders what you have learned through analysis:

**Select.** Communicating what one has learned involves selecting among those analyses that are most important and most useful to an intended audience, as well as selecting a form for displaying that information, such as a graph or table in static or interactive form.

**Polish.** After creating initial versions of data products, research teams often spend time refining or polishing them, by adding or editing titles, labels, and notations and by working with colors and shapes to highlight key points.

**Narrate.** Writing a narrative to accompany the data products involves, at a minimum, pairing a data product with its related research question, describing how best to interpret the data product, and explaining the ways in which the data product helps answer the research question.

For this case study, let's focus on returning to our research questions:

1. What are the patterns of epistemic connections formed by learners as they collaboratively engage in engineering problem-solving tasks within digital internship environments?
2. Is there a difference in learners' epistemic frames based on the condition to which they were assigned?

👉 **Your Turn**

Imagine that you are part of the RescuShell research team responsible for communicating your work to a broader audience. Based on the analyses conducted in Sections 3 and 4, below are three key findings from our epistemic network analyses that are interesting and potentially actionable.

**KEY FINDING 1: Experience reshapes, not just expands, the structure of engineering discourse.**

Across all 48 participants, the most consistently active epistemic connections involved Design.Reasoning paired with Performance.Parameters and Technical.Constraints paired with Data. However, the balance between these clusters shifted dramatically by condition. FirstGame participants (novices) weighted their connections heavily toward Technical.Constraints and Collaboration, while SecondGame participants (relative experts) concentrated their epistemic energy on Design.Reasoning and Performance.Parameters. This is not simply a story of experts knowing more; it reflects a qualitative reorganization of how they think about engineering problems, moving from managing the surface-level structure of the problem space to integrating design logic with performance evaluation criteria. For instructors and internship designers, this suggests that tracking which code pairs become dominant over time is a more informative signal of learning progression than tracking raw participation volume.

**KEY FINDING 2: The epistemic shift from novice to expert is large, statistically robust, and detectable along a single primary dimension.**

Both parametric (Welch two-sample t test: t(45.31) = -6.52, p < 0.001, Cohen's d = 1.88) and non-parametric (Mann-Whitney U: U = 50, p < 0.001, r = 0.86) tests confirm that FirstGame and SecondGame participants occupy meaningfully different positions in ENA's first semantic dimension (MR1), with an effect size in the large range by any standard benchmark. Critically, no significant difference emerged on the second dimension (SVD2), which captured patterns of connection that were stable across conditions. This dimensionality result is itself actionable: the primary axis of difference, anchored by the Design.Reasoning and Performance.Parameters cluster on the expert end, represents a concrete instructional target. Educators wanting to accelerate novice development toward expert-like thinking should design activities that require students to explicitly connect performance criteria to their design justifications, not simply participate more in collaborative discussion.

**KEY FINDING 3: Collaboration serves different epistemic functions at different experience levels, and this has implications for how virtual internships are scaffolded.**

The subtracted network and normalized line weights consistently showed that Collaboration carried higher relative weight in FirstGame than in SecondGame. This is not a failure of expert students to collaborate; rather, it reflects that novices required more explicit, discourse-level coordination to navigate shared problem understanding, while experts moved collaboration into the background and devoted their conversational energy to substantive analytical work. For virtual internship designers, this finding suggests that scaffolding collaboration explicitly (through structured team protocols, shared task checklists, or mentor-prompted coordination prompts) may be most valuable in early phases of the experience and can be reduced as students develop shared domain fluency. Monitoring the Collaboration-to-DesignReasoning ratio in real time via ENA could serve as a formative signal of when a team is ready to transition from coordination mode into deeper analytical engagement.

---

Congratulations! You've completed the Unit 4 Case Study: Intro to ENA. To "turn in" your work, you can click the "Render" icon in the menu bar above. This will create a HTML report in your Files pane that serves as a record of your completed assignment and that can be opened in a browser or shared on the web.

## References
